---
title: "Benchmarks"
output:
   rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{terra, ggplot2, bench, whitebox}
---

```{r knitr_setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  dpi = 150,
  retina = 2
)

requireNamespace("terra", quietly = TRUE)
requireNamespace("whitebox", quietly = TRUE)
requireNamespace("ggplot2", quietly = TRUE)
requireNamespace("tidyr", quietly = TRUE)
requireNamespace("bench", quietly = TRUE)

# Install WhiteboxTools
if (!whitebox::check_whitebox_binary()) {
  whitebox::install_whitebox()
}
whitebox::wbt_verbose(verbose = FALSE)

ggplot2::theme_set(
  ggplot2::theme_bw() +
    ggplot2::theme(axis.text.y = ggplot2::element_text(face = "bold"))
)

```

To evaluate the performance of different R packages for terrain analysis, here three different implementations are benchmarked:

1. **wbw** (WhiteboxWorkflows for R): this package
2. [**whitebox**](https://github.com/opengeos/whiteboxR/) (WhiteboxTools for R): A traditional command-line based approach that writes results to disk
3. [**terra**](https://rspatial.github.io/terra/): R's leading package for raster processing, using an optimized C++ backend

# Hillshade

The traditional hillshading, repeating the procedure from the `terra::shade()` documentation. Each process was run 21 times.

```{r bench_hillshade, message = FALSE, warning = FALSE}
# Load libraries
library(wbw)
library(whitebox)
library(terra)

# Load file
f <- system.file("extdata/dem.tif", package = "wbw")
# Create a tempfile for {whitebox}
tmp <- tempfile(fileext = ".tif")

bench_hillshade <-
  bench::mark(
    wbw = {
      wbw_read_raster(f) |>
        wbw_hillshade(
          azimuth = 270,
          altitude = 40
        )
    },
    whitebox = {
      whitebox::wbt_hillshade(
        f,
        azimuth = 270,
        altitude = 40,
        output = tmp,
        compress_rasters = FALSE
      )
    },
    terra = {
      r <- terra::rast(f)
      s <- terra::terrain(r, "slope", unit = "radians")
      a <- terra::terrain(r, "aspect", unit = "radians")
      terra::shade(slope = s, aspect = a, angle = 40, direction = 270, normalize = TRUE)
    },
    check = FALSE,
    time_unit = "ms",
    iterations = 21L
  )

bench_hillshade

```

All three packages performed adequately, with `wbw` being `r round(as.numeric(bench_hillshade$median[3]) / as.numeric(bench_hillshade$median[1]), 2)` times faster than `terra` and `r round(as.numeric(bench_hillshade$median[2]) / as.numeric(bench_hillshade$median[1]), 2)` times faster than the original `whitebox`. While with the latter, the raster object is not yet loaded into the R session.

```{r hill_bench_plot}
library(ggplot2)

ggplot2::autoplot(bench_hillshade, type = "boxplot") +
  ggplot2::labs(
    title = "Time to compute hillshade from DEM",
    y = "Elapsed time, milliseconds",
    x = NULL
  )

```

One can plot all hillshades next to each other to ensure that they all look alike.

```{r hillshade_plot}

# wbw
wbw_hillshade <-
  wbw_read_raster(f) |>
  wbw_hillshade(
    azimuth = 270,
    altitude = 40
  ) |>
  as_rast()

# whitebox
wbt_hillshade <- terra::rast(tmp)

# terra
r <- terra::rast(f)
s <- terra::terrain(r, "slope", unit = "radians")
a <- terra::terrain(r, "aspect", unit = "radians")
terra_hillshade <-
  terra::shade(
    slope = s,
    aspect = a,
    angle = 40,
    direction = 270,
    normalize = TRUE
  )

# Set up a 1x3 plotting layout
par(mfrow = c(1, 3))
clrs <- grey(0:100/100)

# wbw
plot(wbw_hillshade,
  main = "wbw",
  legend = FALSE,
  col = clrs,
  mar = c(1, 1, 1, 1)
)

# whitebox
plot(wbt_hillshade,
  main = "whitebox",
  legend = FALSE,
  col = clrs,
  mar = c(1, 1, 1, 1)
)

# terra
plot(terra_hillshade,
  main = "terra",
  legend = FALSE,
  col = clrs,
  mar = c(1, 1, 1, 1)
)

```

# Terrain Parameters

When it comes to slope estimation, the difference between packages is not that significant. 

```{r bench_slope}
# Create a tempfile for {whitebox}
tmp <- tempfile(fileext = ".tif")

bench_slope <-
  bench::mark(
    wbw = {
      wbw_read_raster(f) |>
        wbw_slope()
    },
    whitebox = {
      whitebox::wbt_slope(
        f,
        output = tmp,
        compress_rasters = FALSE
      )
    },
    terra = {
      terra::rast(f) |> 
        terra::terrain("slope")
    },
    check = FALSE,
    time_unit = "ms",
    iterations = 21L
  )

bench_slope

```


# Raster Summary Stats and I/O

As a rule of thumb, majority of in-memory operations happens similarly fast in `wbw` compared to `terra`. Mind the difference in performance for the slope estimation when rasters are loaded into R session.

```{r bench_stats, message = FALSE, warning = FALSE}
# Load libraries
library(wbw)
library(terra)

# Load file
f <- system.file("extdata/dem.tif", package = "wbw")
w <- wbw_read_raster(f)
r <- terra::rast(f)

# Conversion raster to matrix
bench::mark(
  wbw = as_matrix(w),
  terra = as.matrix(r, wide = TRUE),
  check = TRUE,
  time_unit = "ms",
  iterations = 21L
)

# Estimating true mean
bench::mark(
  wbw = mean(w),
  terra = global(r, "mean")$mean,
  check = TRUE,
  time_unit = "ms",
  iterations = 21L
)


# Estimating slope
bench::mark(
  wbw = wbw_slope(w),
  terra = terrain(r, "slope"),
  check = FALSE,
  time_unit = "ms",
  iterations = 21L
)

```

However, this happens because `terra` is not actually loading the whole raster into memory, unlike `wbw`. Therefore, the *input* operations are much faster for `terra`, while the *output* operations are more efficient in `wbw`.

```{r bench_io}
# Reading raster
bench::mark(
  wbw = wbw_read_raster(f),
  terra = rast(r),
  check = FALSE,
  time_unit = "ms",
  iterations = 21L
)

# Writing uncompressed raster
tmp_wbw <- tempfile(fileext = ".tif")
tmp_terra <- tempfile(fileext = ".tif")

bench::mark(
  wbw = wbw_write_raster(w, tmp_wbw, compress = FALSE),
  terra = writeRaster(r, tmp_terra, overwrite = TRUE),
  check = FALSE,
  time_unit = "ms",
  iterations = 21L
)

```

```{r cleanup, include = FALSE}
unlink(tmp_wbw)
unlink(tmp_terra)
unlink(tmp)
```


# Session Info

<details>
<summary> Show </summary>
```{r sessioninfo}
sessionInfo()
```
</details>