---
title: "Tutorial 2: Geomorphometric Analysis"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Tutorial 02}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{terra}
---

```{r knitr_setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  dpi = 300,
  retina = 2
)

requireNamespace("terra", quietly = TRUE)
```

> NB! This tutorial is based on the [Whitebox Workflows for Python (WbW) Tutorial 2: Geomorphometric Analysis](https://github.com/jblindsay/jblindsay.github.io/blob/master/WhiteboxTutorials/WbW_tutorials/WbW_tutorial2.ipynb)

# 1. Download sample data

```{r load_layers}
library(wbw)

# Download sample dataset
sample <- 
  wbw_download_sample_data(
    data_set = "Grand_Junction", 
    path = tempdir()
  )

# Load as WhiteboxRaster
dem <- wbw_read_raster(file.path(sample, "DEM.tif"))
dem
```

# 2. Fill missing data

However, there are some missing values presented in the DEM.
```{r plot_one}
library(terra)

dem |> 
  as_rast() |> 
  plot()

```

To fill them, one can use `wbw_fill_missing_data` function:
```{r fill_nodata}
dem_filled <- 
  wbw_fill_missing_data(dem, filter_size = 35, exclude_edge_nodata = TRUE)

dem_filled |> 
  as_rast() |> 
  plot()
```

# 3. Vizualization

To properly vizualize the DEM, let's create a hillshade map of the relief:

```{r hillshade}
dem_hillshade <- wbw_multidirectional_hillshade(dem_filled)

dem_hillshade |> 
  as_rast() |> 
  plot(col = grDevices::gray.colors(10))
```

# 4. Raster summary stats

One can get `WhiteboxRaster` summary statistics by running familiar functions from base R (and `{terra}`):

```{r summary_stats}
# Raw DEM
summary(dem)
wbw_cols(dem) # Number of columns
wbw_rows(dem) # Number of rows
num_cells(dem) # Number of cells
wbw_res(dem) # Raster resolution

# NA-Filled DEM
summary(dem_filled)
wbw_cols(dem_filled) # Number of columns
wbw_rows(dem_filled) # Number of rows
num_cells(dem_filled) # Number of cells
wbw_res(dem_filled) # Raster resolution
```

# 5. Extracting land-surface parameters (LSPs)
Now let's extract some common land-surface parameters (LSPs), the basic building blocks of a geomorphometric analysis.

Slope and aspect are two of the most common LSPs.

```{r slope_and_aspect}
# Slope
slope <- wbw_slope(dem_filled)
aspect <- wbw_aspect(dem_filled)

# Set up a 1x2 plotting layout
par(mfrow = c(1, 2))

# Plot the first raster (Slope)
plot(as_rast(slope),
     main = "Slope",
     legend = FALSE,
     col = terrain.colors(100),
     mar=c(1,1,1,1))

# Plot the second raster (Aspect)
plot(as_rast(aspect),
     main = "Aspect",
     legend = FALSE,
     mar=c(1,1,1,1))

```